---
title: "When thinking from first principles leads to ruin"
author: "Coleman Breen"
format: html
editor: visual
---

Definition of precautionary principle. Non-naive PP. States of ruin.

Biologists and chemists like to think from first principles

Less about epistemology than it is about

Do this in Jupyter notebook.

# What we don't know

Single-stranded DNA. Let's say that If there are five AAATTT "motifs" in a row then it's game over. This virus will have an R0 of ten and we're all fucked.

But there is virtually no way that this can occur naturally. Let's say the fitness function *likes* repetitive elements, but really does not like AATT motifs. Likes having three AAATTT motifs, but no more and no less.

Fitness function is number of AAA TTT

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
